---
title: "Example Slides"
subtitle: "With a super amazing subtitle"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, echo = F, message = F, warning = F, fig.align='center'}
library(tidyverse)
knitr::opts_chunk$set(fig.width = 10, fig.height = 5, fig.align = 'center')
```

# Headline
```{r}
iris %>% 
  ggplot(aes(Sepal.Length, Sepal.Width, col = Species)) +
  geom_point()
```

---
class:middle

## Centering
This slide is now filled from the center due to `class:middle`

???

* Using `???` I can hide notes.
* I can access these from my knitted html document by pressing `P` 
(i.e. going into presentation mode.) 
* Also I can clone the presentation by pressing `C`

---
class:middle

### Line highlighting via #<<

```{r, eval = F}
mpg %>% 
  ggplot(aes(hwy, cty, col = class)) +
  geom_jitter() #<<
# Figure is shown on next slide! (due to --)
```

--

```{r, echo = F}
mpg %>% 
  ggplot(aes(hwy, cty, col = class)) +
  geom_jitter() #<<
```

More information in the [original intro presentation](https://slides.yihui.org/xaringan/#1).